---
title: "Plots for breast cancer"
author: "cyy36@cam.ac.uk"
date: today
bibliography: references.bib
csl: nature.csl
link-citations: true
execute:
  warning: false
  message: false
  echo: false
  cache: false
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    embed-resources: true
    theme:
      light: flatly
      dark: darkly
---

# Cover Page {.unnumbered .unlisted}

<center>
![](https://onlinecareeraccelerators.pace.cam.ac.uk/hubfs/CAM%20DS%20PACE/Professional%20and%20Continuing%20Education_V_reversed%20colour%201.svg)
</center>

### MSt Healthcare Data Science {.unnumbered .unlisted}

#### Authentication of practice

|   |   |
|---|---|
|I confirm that I have fully read and understood the [assignment brief](https://vle.pace.cam.ac.uk/mod/page/view.php?id=2419694) for this module.| **Y** |

#### Details

|   |   |
|---|---|
|Name: Chung Yan|Surname: Yu|
|Submission Date|   |
|Word Count: whole assignment including codes|  |
|Word Count: main body excluding abstract, references and supplementary materials|  |

#### Permission to share your assignment

I do give permission to share my assignment with future MSt participants.

#### University statement of originality

This assignment is the result of my own work and includes nothing which is the outcome of work done in collaboration except as declared in the Preface and specified in the text. It is not substantially the same as any that I have previously submitted for a degree or diploma or other qualification at the University of Cambridge or any other university or similar institution, or that is being concurrently submitted, except as declared in the Preface and specific in the text. I further state that no substantial part of my Portfolio has already been submitted, nor is being concurrently submitted for any such degree, diploma or other qualification at the University of Cambridge or any other university or similar institution except as declared in the Preface and specified in the text.

|   |   |
|---|---|
|I confirm the statement of originality as above| **Y** |


#### Questions for reflection

Self-assessment is an important aspect of feedback literacy, which is, in turn, key to the development of expertise. As you proceed through the MSt Healthcare data science programme, we hope that you will make use of the following prompts to assess your own work on assignments. Specific assignment briefs will likely indicate which of these to address for which assessments, but, in general, we expect you to respond to one or two for each assignment on your course.

For each of the questions, do not spend too long answering – keep it brief. For each question you answer, limit yourself to no more than three items. And please remember, this is optional and developmental: these cover sheets are designed to create space for self-assessment and feedback dialogue, rather than additional assignment workload.

1. Which aspects of this assignment are you most uncertain about and/or would most like to receive feedback on?

- 

2. What elements are you left pondering after this assignment that you would like to discuss further?

- 

3. How have you incorporated feedback from peers and tutors into this assignment?

- 

4. How, and to what extent, have you been able to incorporate feedback on previous course work into this assignment?

- 

5. Using the wording in the rubric, how would you describe the quality of the different aspects of your work?


### Declaration of the use of generative AI {.unnumbered .unlisted}

|   |   |
|---|---|
|Which permitted use of generative AI are you acknowledging?| Semantic search of literature and notes, outline creation, output formatting, informed feedback, code generation |
|Which generative AI tool did you use (name and version)?| Claude Code v2.1.1x (Opus 4.6), M365 CoPilot |
|What did you use the tool for?| Searching for relevant journal publications (via PubMed MCP), collating notes in my Obsidian vault, `.bib` file curation, code debugging+generation |
|How have you used or changed the generative AI's output| My collated notes always stay in point form so that I write out the paragraphs in my own words. Feedback provided by the GenAI models are weighed and assessed before being acted on. |


> The GitHub repository of this project is found at .

# Design rationale

<!--
Max 250 words. Include:
  - Intended audience (who, why)
  - Colour palette choices
  - Labelling and annotation strategy
  - Layout and visual hierarchy
  - Any audience-specific decisions
-->

*[To be written — 250 word limit]*

# Setup

```{r}
#| label: setup
#| include: false

# Key visualisation packages
# (data-pipeline packages — dplyr, survival, GEOquery, cBioPortalData —
# are loaded inside the sourced scripts)
suppressPackageStartupMessages({
  library(ggplot2)
  library(patchwork)
  library(ggrepel)
  library(scales)
  library(viridisLite)
  library(survminer)
  library(forcats)
})

# Source the full analysis pipeline
# The scripts handle their own caching via CSV files in data/processed/,
# so re-running is fast on subsequent renders.
invisible(capture.output(suppressMessages({
  source("scripts/00-fetch-data.R")
  source("scripts/01-immune-scoring.R")
  source("scripts/02-explore.R")
  source("scripts/03-pca-cytotoxic.R")
  source("scripts/04-confirmatory-cox.R")
  source("scripts/05-ispy2.R")
})))
```

# Plot theme and palette

```{r}
#| label: plot-theme
#| echo: true

# ------------------------------------------------------------------
# Palette — Okabe-Ito colourblind-safe set, assigned to narrative
# roles so each colour has one job across all figures.
#
#   METABRIC (discovery cohort)         → blue
#   TCGA (validation cohort)            → bluish green
#   Pembrolizumab arm (Fig 3B only)     → vermillion
#   Tumour / central illustration only  → reddish purple
#   Control / neutral / background      → dark grey
#   Highlight box fill (forest plots)   → cream
#
# Every figure uses at most 2–3 of these at once (Nature-style 3–4
# colour limit). Emphasis on the cytotoxic signature is carried by
# box overlays and shape, not by an extra colour — keeping colour
# reserved for categorical cohort encoding.
# ------------------------------------------------------------------

PALETTE <- list(
  metabric     = "#D55E00",  # Okabe-Ito vermillion
  tcga         = "#0072B2",  # Okabe-Ito blue
  pembro       = "#CC79A7",  # Okabe-Ito reddish purple
  control      = "#4D4D4D",  # dark grey
  neutral      = "#4D4D4D"   # alias
)

# Named vectors for ggplot scale_* helpers
COHORT_COLOURS <- c(
  METABRIC = PALETTE$metabric,
  TCGA     = PALETTE$tcga
)

ARM_COLOURS <- c(
  Pembrolizumab = PALETTE$pembro,
  Control       = PALETTE$control
)

# ------------------------------------------------------------------
# theme_hds() — clean, Tufte-inspired, axis-only, white background
# Designed at base_size=14 for .qmd preview; figures scale up in
# Illustrator for A1 export (minimums: 24pt axis, 22pt ticks).
# ------------------------------------------------------------------
theme_hds <- function(base_size = 14) {
  theme_classic(base_size = base_size) +
    theme(
      # Titles and axes
      plot.title        = element_text(face = "bold", size = base_size * 1.15),
      plot.subtitle     = element_text(colour = "grey40", size = base_size * 0.95),
      axis.title        = element_text(face = "plain", size = base_size),
      axis.text         = element_text(size = base_size * 0.9),
      axis.line         = element_line(colour = "black", linewidth = 0.4),
      axis.ticks        = element_line(colour = "black", linewidth = 0.4),
      # Legend
      legend.title      = element_text(face = "bold", size = base_size * 0.95),
      legend.text       = element_text(size = base_size * 0.9),
      legend.background = element_blank(),
      legend.key        = element_blank(),
      legend.position   = "right",
      # Panels and strip
      panel.background  = element_rect(fill = "white", colour = NA),
      panel.grid        = element_blank(),
      strip.background  = element_blank(),
      strip.text        = element_text(face = "bold", size = base_size * 0.95),
      # Overall
      plot.background   = element_rect(fill = "white", colour = NA)
    )
}

# Convenience scale functions — consistent cohort encoding
scale_colour_cohort <- function(...) scale_colour_manual(values = COHORT_COLOURS, ...)
scale_fill_cohort   <- function(...) scale_fill_manual(values = COHORT_COLOURS, ...)
scale_colour_arm    <- function(...) scale_colour_manual(values = ARM_COLOURS, ...)
scale_fill_arm      <- function(...) scale_fill_manual(values = ARM_COLOURS, ...)

# Diverging gradient for heatmaps: blue ↔ white ↔ vermillion
# (reuses METABRIC and pembro colours as palette anchors)
scale_fill_diverging <- function(limits = c(-1, 1), ...) {
  scale_fill_gradient2(
    low      = PALETTE$tcga,
    mid      = "white",
    high     = PALETTE$metabric,
    midpoint = 0,
    limits   = limits,
    ...
  )
}
```

# Figures

## Figure 1 — Cross-cohort immune cell screen

*Claim: Across 14 immune cell types in METABRIC and TCGA, a protective lymphocyte cluster (cytotoxic cells, B-cells, CD8 T-cells) is the only group consistent across cohorts.*

### 1A. Forest plot: 14 cell types, PAM50-adjusted HRs, both cohorts

```{r}
#| label: fig-1a-forest
#| fig-width: 7
#| fig-height: 8

# --- Prepare data -----------------------------------------------------------
# `comparison` comes from 02-explore.R (long format, both models)
# Saved to data/processed/cox_comparison.csv but the in-memory name is `comparison`
forest_data <- comparison %>%
  filter(model == "PAM50-adjusted") %>%
  mutate(
    cohort = factor(cohort, levels = c("metabric", "tcga"),
                    labels = c("METABRIC", "TCGA"))
  )

# Cell type order: by METABRIC PC1 loading (from script 03, loadings_m)
# This ties Fig 1A visual ordering to Fig 2A (loading plot) — same sequence
cell_order <- names(sort(abs(loadings_m), decreasing = FALSE))  # bottom-up for ggplot
forest_data <- forest_data %>%
  mutate(cell_type = factor(cell_type, levels = cell_order))

# --- Append the fully-adjusted cytotoxic row (f3/g3 PRIMARY models) ---------
# These come from cox_robustness in 04-confirmatory-cox.R. Adding them as a
# new top row lets the forest plot show both the discovery screen (PAM50-
# adjusted, 14 cell types) AND the confirmatory survival result for cytotoxic
# (fully adjusted for PAM50 + age + stage) in one figure. The top row will be
# highlighted in Illustrator to signal its different adjustment level.
primary_row <- cox_robustness %>%
  filter(grepl("PRIMARY", model)) %>%
  transmute(
    cohort    = factor(cohort, levels = c("METABRIC", "TCGA")),
    cell_type = "Cytotoxic (fully adjusted)",
    hr        = HR,
    hr_lower  = HR_lower,
    hr_upper  = HR_upper
  )

# Extend the cell-type ordering so the new row sits at the very top
cell_order_ext <- c(cell_order, "Cytotoxic (fully adjusted)")

forest_data <- bind_rows(
  forest_data %>% select(cohort, cell_type, hr, hr_lower, hr_upper),
  primary_row
) %>%
  mutate(cell_type = factor(cell_type, levels = cell_order_ext))

# Dodge width for per-row cohort offset
dodge_width <- 0.5

# --- Build plot -------------------------------------------------------------
fig_1a <- ggplot(forest_data, aes(x = hr, y = cell_type, colour = cohort)) +
  # Reference line at HR = 1.0
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = PALETTE$neutral, linewidth = 0.4) +
  # 95% CI error bars
  geom_errorbarh(
    aes(xmin = hr_lower, xmax = hr_upper),
    height = 0,
    position = position_dodge(width = dodge_width),
    linewidth = 0.6
  ) +
  # HR point estimates
  geom_point(
    size = 3.5,
    position = position_dodge(width = dodge_width)
  ) +
  # Log-scale x-axis (HR is multiplicative)
  scale_x_continuous(
    trans = "log10",
    breaks = c(0.6, 0.7, 0.8, 0.9, 1.0, 1.1),
    labels = c("0.6", "0.7", "0.8", "0.9", "1.0", "1.1")
  ) +
  # Clean cell type labels: replace underscores with spaces
  scale_y_discrete(labels = function(x) gsub("_", " ", x)) +
  scale_colour_cohort() +
  labs(
    title = "Hazard ratios for METABRIC and TCGA-BRCA cohorts",
    x = "Hazard ratio (per 1 SD, 95% CI)",
    y = "Immune cell populations",
    colour = NULL
  ) +
  theme_hds() +
  theme(
    legend.position = "top",
    legend.justification = "left"
  )

ggsave("fig_1a.svg", fig_1a, width = 8, height = 9, device = "svg")
fig_1a
```

## Figure 2 — Cytotoxic cells anchor the dominant immune axis

*Claim: PCA on all 14 cell types shows cytotoxic cells at the top of PC1 loadings in both cohorts; cytotoxic and B-cells are collinear and either captures the same signal.*

### 2A. Cleveland dot plot — PC1 loadings by cohort

```{r}
#| label: fig-2a-loadings
#| fig-width: 7
#| fig-height: 6

# --- Cleveland dot plot: PC1 loading per cell type, both cohorts -----------
# loadings_m and loadings_t come from 03-pca-cytotoxic.R
# Use absolute values so the comparison is on magnitude (direction is
# arbitrary in PCA — both cohorts already flipped to positive PC1 = hot)
loadings_clev <- data.frame(
  cell_type = names(loadings_m),
  metabric  = abs(loadings_m),
  tcga      = abs(loadings_t[names(loadings_m)])
) %>%
  # Order top-down by METABRIC loading (cytotoxic at the top)
  mutate(
    cell_type = factor(
      cell_type,
      levels = cell_type[order(metabric)]  # ascending → ggplot puts largest at top
    )
  )

# Long format for the dots
loadings_clev_long <- loadings_clev %>%
  tidyr::pivot_longer(
    cols = c(metabric, tcga),
    names_to = "cohort",
    values_to = "loading"
  ) %>%
  mutate(
    cohort = factor(cohort,
                    levels = c("metabric", "tcga"),
                    labels = c("METABRIC", "TCGA"))
  )

cell_order_clev <- levels(loadings_clev$cell_type)  # used by Fig 2B for shared y-ordering

fig_2a <- ggplot(loadings_clev_long, aes(x = loading, y = cell_type)) +
  # Connecting line per cell type — its length shows cross-cohort disagreement
  geom_line(aes(group = cell_type), colour = "grey70", linewidth = 0.6) +
  # Cohort dots
  geom_point(aes(colour = cohort), size = 3.5) +
  # Cell type labels back on the left; heatmap (2B) sits to the right in
  # the Illustrator layout and its own y-axis provides the pairing
  scale_y_discrete(labels = function(x) gsub("_", " ", x)) +
  scale_colour_cohort() +
  labs(
    title = "Principal component contribution of immune populations\nfor METABRIC & TCGA cohorts",
    x = "Absolute PC1 loading",
    y = "Immune cell populations",
    colour = NULL,
    subtitle = sprintf("PC1 explains %.1f%% of variance (METABRIC)", var_pc1)
  ) +
  theme_hds() +
  theme(
    legend.position = "top",
    legend.justification = "left",
    # Horizontal y-axis title (non-default) — sits left of tick labels
    axis.title.y = element_text(angle = 0, vjust = 0.5, hjust = 0.5,
                                margin = margin(r = 10))
  )

ggsave("fig_2a.svg", fig_2a, width = 7, height = 6, device = "svg")
fig_2a
```

### 2B. Collinearity heatmap (METABRIC, 14 × 14)

```{r}
#| label: fig-2b-heatmap
#| fig-width: 7
#| fig-height: 6

# --- METABRIC pairwise correlation between cell type scores --------------
score_cols_2b <- setdiff(
  colnames(scores_metabric),
  c("patient_id", "cohort", "T_cells")
)

corr_m <- cor(surv_metabric[, score_cols_2b], use = "pairwise.complete.obs")

# Reorder the matrix rows/columns to match the plotting order (Fig 2A)
# BEFORE blanking the upper triangle, so the blanked half aligns with the
# visual layout. Without this, the triangle gets computed on alphabetical
# order and plotted on loading order → random missing rows/columns.
corr_m <- corr_m[cell_order_clev, cell_order_clev]

# Blank the upper triangle — symmetric matrix, upper half is redundant.
# Keep diagonal (self-correlations = 1) as a visual anchor.
corr_m[upper.tri(corr_m)] <- NA

# Long format — factor levels match the matrix ordering
corr_long <- as.data.frame(corr_m) %>%
  tibble::rownames_to_column("cell_a") %>%
  tidyr::pivot_longer(-cell_a, names_to = "cell_b", values_to = "r") %>%
  filter(!is.na(r)) %>%  # drop the blanked upper triangle
  mutate(
    cell_a = factor(cell_a, levels = cell_order_clev),
    cell_b = factor(cell_b, levels = cell_order_clev)
  )

fig_2b <- ggplot(corr_long, aes(x = cell_a, y = cell_b, fill = r)) +
  geom_tile(colour = "white", linewidth = 0.3) +
  geom_text(
    aes(label = sprintf("%.2f", r)),
    size = 2.5,
    colour = ifelse(abs(corr_long$r) > 0.6, "white", "black")
  ) +
  scale_fill_diverging(limits = c(-1, 1), name = "Pearson r") +
  scale_x_discrete(labels = function(x) gsub("_", " ", x)) +
  scale_y_discrete(
    labels = function(x) gsub("_", " ", x),
    position = "right"  # sit on the right so the labels face outward,
                        # away from Fig 2A in the Illustrator layout
  ) +
  labs(
    title = "Immune population score collinearity (METABRIC)",
    x = "Immune cell populations",
    y = NULL,
    subtitle = "Cell populations ordered by METABRIC PC1 loading (Fig 2A)"
  ) +
  theme_hds() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.border = element_rect(colour = NA, fill = NA)
  )

ggsave("fig_2b.svg", fig_2b, width = 7, height = 6, device = "svg")
fig_2b
```

### 2C. Progressive adjustment forest (cytotoxic HR across models)

```{r}
#| label: fig-2c-progressive
#| fig-width: 8
#| fig-height: 5

# --- Prepare data: the cox_robustness table from 04-confirmatory-cox.R ---
# Drop METABRIC sensitivity-only rows (grade, NPI) — those are internal
# checks, not part of the core comparison with TCGA
adjustment_data <- cox_robustness %>%
  filter(!grepl("Grade|NPI", model)) %>%
  mutate(
    cohort = factor(cohort, levels = c("METABRIC", "TCGA")),
    # Uniform short labels so both cohorts stack cleanly top-to-bottom
    model_label = case_when(
      grepl("Unadjusted", model)    ~ "Unadjusted",
      grepl("^\\+PAM50$", model)    ~ "+ PAM50",
      grepl("\\+Age$", model)        ~ "+ Age",
      grepl("Stage \\[PRIMARY\\]", model) ~ "+ Stage\n[PRIMARY]",
      TRUE ~ model
    ),
    model_label = factor(
      model_label,
      levels = rev(c("Unadjusted", "+ PAM50", "+ Age", "+ Stage\n[PRIMARY]"))
    )
  )

fig_2c <- ggplot(adjustment_data, aes(x = HR, y = model_label, colour = cohort)) +
  # Reference line at HR = 1
  geom_vline(xintercept = 1, linetype = "dashed",
             colour = PALETTE$neutral, linewidth = 0.4) +
  # 95% CI bars
  geom_errorbarh(
    aes(xmin = HR_lower, xmax = HR_upper),
    height = 0, linewidth = 0.6
  ) +
  # HR estimates
  geom_point(size = 3.5) +
  # Side-by-side panels, not stacked
  facet_wrap(~ cohort, ncol = 2) +
  scale_x_continuous(
    trans = "log10",
    breaks = c(0.7, 0.8, 0.9, 1.0, 1.1)
  ) +
  scale_colour_cohort(guide = "none") +
  labs(
    x = "Cytotoxic cells HR (per 1 SD)",
    y = NULL
  ) +
  theme_hds()

ggsave("fig_2c.svg", fig_2c, width = 8, height = 5, device = "svg")
fig_2c
```

## Figure 3 — Clinical translation

*Claim: Higher cytotoxic infiltration is associated with longer survival (METABRIC) and improved pCR in the I-SPY2 pembrolizumab arm.*

### 3A. Kaplan-Meier — cytotoxic high vs low (METABRIC)

```{r}
#| label: fig-3a-km
#| fig-width: 7
#| fig-height: 5

# --- Median split on cytotoxic score, METABRIC ---------------------------
surv_m_km <- surv_metabric %>%
  mutate(
    cyto_group = factor(
      ifelse(Cytotoxic_cells >= median(Cytotoxic_cells),
             "High cytotoxic", "Low cytotoxic"),
      levels = c("Low cytotoxic", "High cytotoxic")
    )
  )

fit_km_m <- survfit(Surv(os_months, os_event) ~ cyto_group, data = surv_m_km)

# survminer::ggsurvplot returns a list with $plot — extract it for theming
km_plot <- ggsurvplot(
  fit_km_m,
  data         = surv_m_km,
  conf.int     = TRUE,
  palette      = c(PALETTE$neutral, PALETTE$metabric),
  pval         = TRUE,
  pval.coord   = c(150, 0.85),
  xlab         = "Time (months)",
  ylab         = "Overall survival probability",
  legend.title = "",
  legend.labs  = c("Low cytotoxic", "High cytotoxic"),
  ggtheme      = theme_hds()
)

# Strip the fill-box from the legend key (leaves only the coloured line)
fig_3a <- km_plot$plot +
  guides(
    colour = guide_legend(override.aes = list(fill = NA)),
    fill   = "none"
  ) +
  labs(title = "Overall survival by cytotoxic infiltration (METABRIC)")

ggsave("fig_3a.svg", fig_3a, width = 7, height = 5, device = "svg")
fig_3a
```

### 3B. I-SPY2 logistic curves (pembro vs control)

```{r}
#| label: fig-3b-logistic
#| fig-width: 7
#| fig-height: 5

# --- Prepare data: stack pembro + control with arm label and pooled cyto_z --
# `pooled` is in memory from 05-ispy2.R (pembro + control, scaled once)
ispy_plot_data <- pooled %>%
  mutate(
    arm_label = factor(
      ifelse(is_pembro == 1, "Pembrolizumab", "Control"),
      levels = c("Pembrolizumab", "Control")
    )
  )

fig_3b <- ggplot(ispy_plot_data,
                 aes(x = Cytotoxic_cells_z, y = pcr, colour = arm_label, fill = arm_label)) +
  # Rug marks for individual patients (top for pCR=1, bottom for pCR=0)
  geom_point(
    position = position_jitter(height = 0.03, width = 0),
    alpha = 0.5, size = 1.2
  ) +
  # Logistic regression curve per arm with 95% CI ribbon
  geom_smooth(
    method = "glm",
    method.args = list(family = "binomial"),
    se = TRUE, alpha = 0.15, linewidth = 0.8
  ) +
  scale_colour_arm(
    guide = guide_legend(override.aes = list(fill = NA))  # strip fill from key
  ) +
  scale_fill_arm(guide = "none") +  # hide fill from legend — only line shows
  scale_y_continuous(
    breaks = c(0, 0.25, 0.5, 0.75, 1),
    labels = c("0", "0.25", "0.50", "0.75", "1"),
    limits = c(-0.05, 1.05)
  ) +
  labs(
    title = "Pathological complete response in I-SPY2 by cytotoxic score",
    x = "Cytotoxic immune cell population score (normalised & pooled)",
    y = "Pathological complete response",
    colour = NULL,
    fill = NULL
  ) +
  theme_hds() +
  theme(
    legend.position = "top",
    legend.justification = "left"
  )

ggsave("fig_3b.svg", fig_3b, width = 8, height = 5, device = "svg")
fig_3b
```

# Supplementary figures

## Cox assumption tests

### Schoenfeld residuals (proportional hazards)

Each panel plots the scaled Schoenfeld residual for one model term against
transformed survival time. A flat trend indicates the proportional hazards
assumption holds.

```{r}
#| label: fig-sup-schoenfeld
#| fig-width: 10
#| fig-height: 8

# Manually extract Schoenfeld residual data from cox.zph() objects
# (ph_m = METABRIC f3, ph_t = TCGA g3 from 04-confirmatory-cox.R)
extract_schoenfeld <- function(ph_obj, cohort_label) {
  terms <- colnames(ph_obj$y)
  pvals <- ph_obj$table[terms, "p"]
  # Long-format data frame: one row per (term, time) residual point
  do.call(rbind, lapply(seq_along(terms), function(i) {
    data.frame(
      cohort  = cohort_label,
      term    = terms[i],
      term_label = sprintf("%s (p=%.3f)", terms[i], pvals[i]),
      time    = ph_obj$x,
      resid   = ph_obj$y[, i],
      stringsAsFactors = FALSE
    )
  }))
}

sch_df <- bind_rows(
  extract_schoenfeld(ph_m, "METABRIC"),
  extract_schoenfeld(ph_t, "TCGA")
) %>%
  mutate(cohort = factor(cohort, levels = c("METABRIC", "TCGA")))

fig_sup_sch <- ggplot(sch_df, aes(x = time, y = resid, colour = cohort, fill = cohort)) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = PALETTE$neutral, linewidth = 0.4) +
  geom_point(alpha = 0.3, size = 0.6) +
  geom_smooth(method = "loess", se = TRUE, alpha = 0.15, linewidth = 0.6) +
  facet_grid(term ~ cohort, scales = "free", labeller = label_wrap_gen(width = 15)) +
  scale_colour_cohort(guide = "none") +
  scale_fill_cohort(guide = "none") +
  labs(
    x = "Transformed survival time",
    y = "Scaled Schoenfeld residual"
  ) +
  theme_hds(base_size = 10)

fig_sup_sch
```

### Martingale residuals (linearity)

Martingale residuals from a null Cox model (confounders only, no cytotoxic)
plotted against the raw cytotoxic score. A flat or straight-line trend
means the linear assumption is appropriate.

```{r}
#| label: fig-sup-martingale
#| fig-width: 10
#| fig-height: 4

# METABRIC: mart_resid_m from 04, surv_m_stg contains the raw cytotoxic column
mart_df_m <- data.frame(
  cohort    = "METABRIC",
  cytotoxic = surv_m_stg$Cytotoxic_cells,
  resid     = mart_resid_m
)
mart_df_t <- data.frame(
  cohort    = "TCGA",
  cytotoxic = surv_t_stg$Cytotoxic_cells,
  resid     = mart_resid_t
)
mart_df <- bind_rows(mart_df_m, mart_df_t) %>%
  mutate(cohort = factor(cohort, levels = c("METABRIC", "TCGA")))

fig_sup_mart <- ggplot(mart_df, aes(x = cytotoxic, y = resid, colour = cohort, fill = cohort)) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = PALETTE$neutral, linewidth = 0.4) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "loess", alpha = 0.15, linewidth = 0.6) +
  facet_wrap(~ cohort, scales = "free_x") +
  scale_colour_cohort(guide = "none") +
  scale_fill_cohort(guide = "none") +
  labs(
    x = "Cytotoxic score (raw)",
    y = "Martingale residual"
  ) +
  theme_hds()

fig_sup_mart
```

## PCA variance structure

### Scree plot

PC1 dominates the immune axis with ~38% variance, justifying the
single-component focus in Figure 2.

```{r}
#| label: fig-sup-scree
#| fig-width: 6
#| fig-height: 4

# pca_m comes from 03-pca-cytotoxic.R
scree_df <- data.frame(
  pc       = factor(paste0("PC", seq_along(summary(pca_m)$importance[2, ])),
                    levels = paste0("PC", seq_along(summary(pca_m)$importance[2, ]))),
  variance = summary(pca_m)$importance[2, ] * 100
)

fig_sup_scree <- ggplot(scree_df, aes(x = pc, y = variance, group = 1)) +
  geom_hline(
    yintercept = 100 / nrow(scree_df),
    linetype = "dashed", colour = PALETTE$neutral, linewidth = 0.4
  ) +
  geom_line(colour = PALETTE$metabric, linewidth = 0.6) +
  geom_point(colour = PALETTE$metabric, size = 2.5) +
  labs(
    x = NULL,
    y = "Variance explained (%)",
    caption = "Dashed line: equal variance baseline (100% / 14 PCs)"
  ) +
  theme_hds() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

fig_sup_scree
```

# References

::: {#refs}
:::


